Musical elements in the discrete-time representation of sound

### 2017-07-08 article/acm/body.tex
finished review of section 2

\section{Variations in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}

The basic digital music note defined in section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre. This is a useful and paradigmatic model, but it does not exhaust all the aspects of a musical note. First of all, characteristics of the note modify along the note itself~\cite{Chowning}. For example, a 3 s piano note has intensity with abrupt rise at the beginning and progressive decay, has spectrum variations with harmonics decaying and some others emerging along time. These variations are not mandatory, but they are used in sound synthesis for music because they reflect how sounds appear in nature. This is considered to be so true that there is a rule of thumb: to make a sound that incites interest by itself, do internal variations on it~\cite{Roederer}.
To explore all the ways in which variations occur within a note is out of the scope of any work, given the sensibility of the human ear and the complexity of human sound cognition. In the following, primary resources are presented to produce variations in the basic note. It is worthwhile to recall that all the relations in this and other sections are implemented in Python and published as the public domain \massa\ toolbox. The musical pieces \emph{Transita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} were made to validate and illustrate concepts of this section. The code that synthesizes these pieces are also part of the toolbox\cite{MASSA}.

\subsection{Lookup Table}\label{subsec:lookup}

The \emph{Lookup Table} (or simply LUT) is an array for indexed operations which substitutes continuous and repetitive calculation. It is used to reduce computational complexity and employ functions without direct calculation, as data sampled from nature.
In music its usage transcends these applications, as it simplifies many operations and makes it possible to use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.

\begin{figure*}
    \centering
        \includegraphics[width=.7\textwidth]{figures/lut_}
    \caption{Consultation in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.
	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_a} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{l}$ and $f_a$ is the sampling rate.}
        \label{fig:lut}                                                                                                            
\end{figure*}

@@

The expression defining the variable $\gamma_i$ can be understood as $f_s$ being added to $i$ at each second.
If $i$ is divided by the sample frequency, $\frac{i}{f_a}$
is incremented in $1$ at each second. Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_a}$, which covers the period in each second. Finally, with frequency $f$ it results in $i . f \frac{\widetilde{\Lambda}}{f_a}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e. the resulting sequence presents the fundamental frequency $f$.

@@

\subsection{Incremental Variations of Frequency and Intensity}\label{subsec:vars}

As stated by the Weber and Fechner~\cite{Weber-Fechner} law, the human perception has a logarithmic relation with the stimulus. That is to say, the exponential progression of a stimulus is perceived as linear.
For didactic reasons, and given its use in AM and FM synthesis (subsection~\ref{subsec:tvaf}), linear variation is discussed first.

@@

where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ is the LUT increment between two samples given the sound frequency of the first sample. Therefore, it is handy to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$. Equations~\ref{freqLinear},~\ref{indiceLinear} and~\ref{serieAmostralLin} are related with the linear progression of the frequency. As stated above, the frequency progression \emph{perceived} as linear follows an exponential progression, i.e. a geometric progression of frequency is perceived as an arithmetic progression of pitch. It is possible to write: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ where  $n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ is the number of octaves between $f_0$ and $f_{\Lambda-1}$.
Therefore, $f_i=f_0 . 2^{\frac{i}{\Lambda-1}\log_2\frac{f_{\Lambda-1}}{f_0}}=
 f_0 . 2^{\log_2\left ( \frac{f_{\Lambda-1}}{f_0} \right )^{\frac{i}{\Lambda-1}}}=
 f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}$. Accordingly, the equations for linear pitch transition are:

@@

The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible raise it to a power in such a way that the beginning of the transition will be smoother or steeper. This procedure is useful for energy variations with the purpose of changing the volume\footnote{The volume (psychophysical quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy. Wave energy is easy to modify (see Equation~\ref{eq:potencia}) and can be varied in many ways. The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number. The increase of energy without amplitude variation is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}}. It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of the transition.

Thus, for amplitude variations:

\begin{equation}\label{seqAmp}
\begin{split}
 \{a_i\}_0^{\Lambda-1}= & \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \\ = & \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1} \text{ (with } a_0=1 \text{)}
\end{split}
\end{equation}
and
\begin{equation}\label{transAmp}
\begin{split}
 T_i^{'}=T_i \odot A_i = & \{t_i . a_i\}_0^{\Lambda-1} \\ = & \left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
\end{split}
\end{equation}

It is often convenient to have $a_0=1$ to start a new sequence with the original amplitude and then progressively change it.
If $\alpha=1$, the amplitude variation follows the geometric progression that defines the linear variation of volume. Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by equation~\ref{eq:ampVol}.

Special attention should be dedicated while considering $a=0$.
In equation~\ref{seqAmp}, $a_0=0$ results in a division by zero and if $a_{\Lambda-1}=0$, there will be multiplication by zero. Both cases make the procedure useless, once no number different from zero can be represented as a ratio in relation to zero. It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum volume for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$). A linear fade can be used to reach zero amplitude. Another common solution is the use of the quartic polynomial term $x^4$, as it reaches zero without these difficulties and gets reasonably close to the curve with $\alpha=1$ as it withdraws from zero~\cite{Cook}.

@@

The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in subsection~\ref{subsec:tvaf}. A non-oscillatory exploration of these variations is in the music piece \emph{Transita para metro}, whose code is online as part of the \massa\ toolbox\cite{MASSA}.

\subsection{Application of Digital Filters}\label{subsec:filtros}

This subsection is limited to a description of sequences processing, by convolution and differential equations, and immediate applications, as its complexity escapes the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations. The reader is welcome to visit the bibliography for further discussion~\cite{Openheim,smith}.}. Filter applications can be part of the synthesis process or made subsequently as part of processes commonly referred to as ``sound treatment''.

\subsubsection{Convolution and finite impulse response (FIR) filters}

\begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{figures/convolucao_}
     \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
         \label{fig:conv}
\end{figure*}

Filters applied by means of convolution are known by the acronym FIR (Finite Impulse Response) and are characterized by having a finite sample representation. This sample representation is called `impulse response' $\{h_i\}$. FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}. For the purposes of this work, convolution is defined as:

@@

With this procedure it is possible to apply reverberators, equalizers, \emph{delays}, to name a few of a variety of other filters for sound processing and to obtain musical/artistic effects.

The impulse response can be provided by physical measures or by pure synthesis. An impulse response for a reverberation application, for example, can be obtained by sound recording of the environment when someone triggers a snap which resembles an impulse, or obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response. Both are impulse responses which, properly convoluted with the sound sequence, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}. The inverse Fourier transform of an even and real envelope is an impulse response of a FIR filter. Convoluted with a sound, it performs the frequency filtering specified by the envelope. The greater the number of samples, the higher the envelope resolution and the computational complexity, which should often be weighted, for convolution is expensive.

An important property is the time shift caused by convolution with a shifted impulse. Despite being computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound. Figure~\ref{fig:delays} shows the shift caused by convolution with an impulse. Depending on the intensity of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies). In the latter case, the process resembles granular synthesis, reverbs and equalization.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/delays_}
    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Represented in increasing order of its pulse density. The vertical axis is related to amplitude although one should keep in mind that each plot has two or three displaced signals.}
        \label{fig:delays}
\end{figure*}

\subsubsection{Infinite impulse response (IIR) filters}

This class of filters, known by the acronym IIR, is characterized by having an infinite time representation, i.e.\ the impulse response does not converge to zero. Its application is usually made by the following equation:

@@

In most cases the variables may be normalized: $a_j'=\frac{a_j}{b_0}$ and $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
Equation~\ref{eq:diferencas} is called `difference equation' because the resulting samples $\left\{t_i'\right\}$ are given by differences between original samples $\{t_i\}$ and previous resulting ones $\left\{t_{i-k}'\right\}$.

There are many methods and tools to obtain IIR filters. The text below lists a selection for didactic purposes and as a reference. They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}. For filters of simple order, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
For band-pass and band-reject (or 'notch') filters, this attenuation has two specifications: $f_c$ (in this case, the `center frequency') and bandwidth $bw$. In both frequencies $f_c \pm bw$ there is an attenuation of $\approx 0.707$ of the original amplitude.
There is sound amplification in band-pass and band-reject filters when the cutoff frequency is low and the band width is large enough. In trebles, those filters present only a deviation of the expected profile, expanding the envelope to the bass.

For filters with other frequency responses, it is possible to apply them successively. Another possibility is to use a biquad 'filter receipt'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e. its first direct form consists of two quadratic polynomials forming the fraction: $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 . z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
Both alternatives are explored by ~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.

@@

\begin{enumerate}
  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of Figure~\ref{fig:iir}. The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs. The coefficients $a_0$ and $b_1$ of the IIR filter are given by the intermediate variable $x \in [e^{-\pi},1]$:

\begin{equation}\label{eq:passa-baixas}
 \begin{split}
 x & =e^{-2\pi f_c} \\
 a_0 & =  1-x \\
 b_1 & =  x
 \end{split}
\end{equation}

  \item High-pass filter with a simple pole, module of its frequency responses at the upper right corner of Figure~\ref{fig:iir}. The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of the intermediate variable $x \in [e^{-\pi},1]$:

\begin{equation}\label{eq:passa-altas}
 \begin{split}
 x & =e^{-2\pi f_c} \\
 a_0 & =  \frac{x+1}{2} \\
 a_1 & =  -\frac{x+1}{2} \\
 b_1 & =  x
 \end{split}
\end{equation}

%\item Passa-banda
%\item Rejeita-banda

\item Notch filter. This filter is parametrized by a center frequency $f_c$ and bandwidth $bw$, both given as fraction of $f_s$, therefore $f,\; bw \in (0,0.5)$. Both frequencies $f_c \pm bw$ have $\approx 0.707$ of the amplitude, i.e. attenuation of $3dB$. The auxiliary variables $K$ and $R$ are defined as:

\begin{equation}\label{eq:varAux}
 \begin{split}
  R & = 1 - 3bw \\
  K & = \frac{1-2R\cos(2\pi f_c) + R^2}{2 - 2 \cos (2 \pi f_c)}
 \end{split}
\end{equation}

The band-pass filter in the lower left corner of Figure~\ref{fig:iir} has the following coefficients:

\begin{equation}\label{eq:passa-banda}
 \begin{split}
 a_0 & =  1 - K \\
 a_1 & =  2(K-R)\cos (2\pi f_c) \\
 a_2 & =  R^2-K \\
 b_1 & =  2R \cos (2\pi f_c) \\
 b_2 & =  -R^2
 \end{split}
\end{equation}

The coefficients of band-reject filter, depicted in the lower right of Figure~\ref{fig:iir}, are:

\begin{equation}\label{eq:rejeita-banda}
 \begin{split}
 a_0 & =  K \\
 a_1 & =  -2K\cos (2\pi f_c) \\
 a_2 & =  K \\
 b_1 & =  2R \cos (2\pi f_c) \\
 b_2 & =  -R^2
\end{split}
\end{equation}

%\item Biquad: pela especificação de uma frequência central, da qualidade
%e da intensidade do filtro, este filtro é simples e usual para áudio,
%permitindo ajustes mais finos. Diversas receitas podem ser encontradas
%na literatura, recomendamos especialmente as diferentes especificações
%em ~\ref{musicDSP} e ~\ref{dspguide}.

\end{enumerate}

\subsection{Noise}\label{subsec:ruidos}

Sounds without an explicit pitch are generally called noise~\cite{Lacerda}. They are important musical sounds, as noise is present in piano notes, violin, etc. Furthermore, the majority of percussion instruments does not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}. In electronic music, including electro-acoustic and dance genres, noise has diverse uses and frequently characterizes the music style~\cite{Cook}.

@@

Some noises with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors.
Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic sequence side-by-side. All five noises were generated with the same phase for each component, making it possible to observe the contributions of different parts of the spectrum.

\begin{itemize}

 \item The white note has its name because its energy is distributed equally among all frequencies. It is possible to obtain white noise with the inverse transform of the following coefficients:

\begin{equation}\label{eq:branco}
 \begin{split}
 c_0 & =0 \;\text{,\quad to avoid bias} \\
 c_i & =e^{j.x}\;,\; x \; \text{random} \; \in \; [0,2\pi]\;,\; i \; \in \; \left[1, \, \frac{\Lambda}{2}-1\right] \\
 c_{\Lambda/2} & = 1 \; \text{, \; if $\Lambda$ even}\\ 
 c_i & = c_{\Lambda - i}^*\;,\;\; \text{for}\;  i \; > \;  \frac{\Lambda}{2}
 \end{split}
\end{equation}

@@

 \item The gray noise is defined as a white noise subject to one of the ISO-audible curves. Those curves are obtained by experiments and are imperative to obtain $\alpha_i$. An implementation of ISO 226, which is the last revision of these curves, is in the \massa\ toolbox~\cite{MASSA}.

\end{itemize}

>>>>

\section{Variation in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
The basic digital music note defined in section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre. This is a useful and paradigmatic model, but it does not exhaust all the aspects of a musical note. First of all, characteristics of the note change along the note itself~\cite{Chowning}. For example, a $3s$ piano note has intensity with abrupt rise at the beginning and progressive decay, has spectrum variations with harmonics decaying and some others emerging along time. These variations are not mandatory, but they are used in sound synthesis for music because they reflect how sounds appear in nature. This is considered so important that there is a rule of thumb: to make a sound that incites interest by itself, arrange internal variations on it~\cite{Roederer}.
To explore all the ways in which variations occur within a note is out of the scope of any work, given the sensibility of the human ear and the complexity of human sound cognition. In the following, primary resources are presented to produce variations in the basic note. It is worthwhile to recall that all the relations in this and other sections are implemented in Python and published in the public domain \massa\ toolbox. The musical pieces \emph{Transit para meter}, \emph{Vibrates and shakes}, \emph{Tremolos, vibratos and the frequency}, \emph{Little train of impulsive hillbillies}, \emph{Noisy band}, \emph{Bela rugosi}, \emph{Children's chorus}, \emph{ADa and SaRa} were made to validate and illustrate concepts of this section. The code that synthesizes these pieces is also part of the toolbox\cite{MASSA}.

\subsection{Lookup Table}\label{subsec:lookup}
The \emph{Lookup Table} (or simply LUT) is an array for indexed operations which substitutes continuous and repetitive calculation. It is used to reduce computational complexity and for employing functions without calculating them directly, e.g. from data sampled from nature.
In music its usage transcends these applications: it simplifies many operations and enables the use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.

\begin{figure*}
    \centering
        \includegraphics[width=.7\textwidth]{figures/lut_}
    \caption{Search in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.
	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_a} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{l}$ and $f_a$ is the sampling rate.}
        \label{fig:lut}
\end{figure*}

@@

The expression defining the variable $\gamma_i$ can be understood as $f_s$ being added to $i$ at each second.
If $i$ is divided by the sample frequency, $\frac{i}{f_a}$
is incremented by $1$ at each second. Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_a}$, which covers the period in one second. Finally, with frequency $f$ it results in $i . f \frac{\widetilde{\Lambda}}{f_a}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e. the resulting sequence presents the fundamental frequency $f$.

@@

\subsection{Incremental Variations of Frequency and Intensity}\label{subsec:vars}
As stated by the Weber and Fechner~\cite{Weber-Fechner} law, the human perception has a logarithmic relation with the stimulus. That is to say, the exponential progression of a stimulus is perceived as linear.
For didactic reasons, and given its use in AM and FM synthesis (subsection~\ref{subsec:tvaf}), linear variation is discussed first.

@@

\noindent where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ is the LUT increment between two samples given the sound frequency of the first sample. Therefore, it is handy to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$. Equations~\ref{freqLinear},~\ref{indiceLinear} and~\ref{serieAmostralLin} are related with the linear progression of the frequency. As stated above, the frequency progression \emph{perceived} as linear follows an exponential progression, i.e. a geometric progression of frequency is perceived as an arithmetic progression of pitch. It is possible to write: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ where $n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ is the number of octaves between $f_0$ and $f_{\Lambda-1}$.
Therefore, $f_i=f_0 . 2^{\frac{i}{\Lambda-1}\log_2\frac{f_{\Lambda-1}}{f_0}}=
 f_0 . 2^{\log_2\left ( \frac{f_{\Lambda-1}}{f_0} \right )^{\frac{i}{\Lambda-1}}}=
 f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}$. Accordingly, the equations for linear pitch transition are:

@@

The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible to raise it to a power in such a way that the beginning of the transition will be smoother or steeper. This procedure is useful for energy variations with the purpose of changing the volume\footnote{The volume (psychophysical quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy. Wave energy is easy to modify (see Equation~\ref{eq:potencia}) and can be varied in many ways. The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number. The increase of energy without amplitude variation is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}.}. It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of the transition.

Thus, for amplitude variations:
\begin{equation}\label{seqAmp}
\begin{split}
 \{a_i\}_0^{\Lambda-1}= & \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \\ = & \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1} \text{ (with } a_0=1 \text{)}
\end{split}
\end{equation}
and
\begin{equation}\label{transAmp}
\begin{split}
 T_i^{'}=T_i \odot A_i = & \{t_i . a_i\}_0^{\Lambda-1} \\ = & \left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
\end{split}
\end{equation}

It is often convenient to have $a_0=1$ to start a new sequence with the original amplitude and then progressively change it.
If $\alpha=1$, the amplitude variation follows the geometric progression that implies the linear variation of volume. Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by equation~\ref{eq:ampVol}.

Special attention should be dedicated while considering $a=0$.
In equation~\ref{seqAmp}, $a_0=0$ results in a division by zero and if $a_{\Lambda-1}=0$, there will be multiplication by zero. Both cases make the procedure useless, once a ratio of any number in relation to zero is not well defined for our purposes. It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum volume for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$). A linear fade can be used then to reach zero amplitude, if needed. Another common solution is the use of the quartic polynomial term $x^4$, as it reaches zero without these difficulties and gets reasonably close to the curve with $\alpha=1$ as it departs from zero~\cite{Cook}.

@@

The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in subsection~\ref{subsec:tvaf}. A non-oscillatory exploration of these variations is in the music piece \emph{Transit para meter}, whose code is online as part of the \massa\ toolbox\cite{MASSA}.

\subsection{Application of Digital Filters}\label{subsec:filtros}
This subsection is limited to a description of sequences processing, by convolution and difference equations, and immediate applications, as a thorough discussion of filtering is beyond the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations~\cite{Openheim,smith}.}. Filter applications can be part of the synthesis process or made subsequently as part of processes commonly referred to as ``acoustic/sound treatment''.

\subsubsection{Convolution and finite impulse response (FIR) filters}
\begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{figures/convolucao_}
     \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
         \label{fig:conv}
\end{figure*}

Filters applied by means of convolution are known by the acronym FIR (Finite Impulse Response) and are characterized by having a finite sample representation. This sample representation is called `impulse response' $\{h_i\}$. FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both the sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}. For the purposes of this work, convolution is defined as:

@@

With this procedure it is possible to achieve reverberators, equalizers, \emph{delays}, to name a few of a variety of other filters for sound processing used to obtain musical/artistic effects.

The impulse response can be provided by physical measures or by pure synthesis. An impulse response for a reverberation application, for example, can be obtained by sound recording of the environment when someone triggers a click which resembles an impulse, or obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response. Both are impulse responses which, properly convoluted with the sound sequence, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}. The inverse Fourier transform of an even and real envelope is an impulse response of a FIR filter. Convoluted with a sound, it performs the frequency filtering specified by the envelope. The greater the number of samples, the higher the envelope resolution and the computational complexity, which should often be weighted, for convolution is expensive.

An important property is the time shift caused by convolution with a shifted impulse. Despite being computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound. Figure~\ref{fig:delays} shows the shift caused by convolution with an impulse. Depending on the density of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies). In the latter case, the process resembles granular synthesis, reverberation and equalization.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/delays_}
    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Represented in increasing order of its pulse density. The vertical axis is related to amplitude although one should keep in mind that each subplot has two or three displaced signals.}
        \label{fig:delays}
\end{figure*}

\subsubsection{Infinite impulse response (IIR) filters}
This class of filters, known by the acronym IIR, is characterized by having an infinite time representation, i.e.\ the impulse response does not converge to zero. Its application is usually made by the following equation:

@@

In most cases the variables may be normalized: $a_j'=\frac{a_j}{b_0}$ and $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
Equation~\ref{eq:diferencas} is called `difference equation' because the resulting samples $\left\{t_i'\right\}$ are given by differences between original samples $\{t_i\}$ and previous ones in the resulting signal $\left\{t_{i-k}'\right\}$.

There are many methods and tools to obtain IIR filters. The text below lists a selection for didactic purposes and as a reference. They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}. For filters of simple order, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
For band-pass and band-reject (or 'notch') filters, this attenuation has two specifications: $f_c$ (in this case, the `center frequency') and bandwidth $bw$. In both frequencies $f_c \pm bw$ there is an attenuation of $-3dB \approx 0.707$ of the original amplitude.
There is sound amplification in band-pass and band-reject filters when the cutoff frequency is low and the band width is large enough. In trebles, those filters present only a deviation of the expected profile, expanding the envelope to the bass.

It is possible to apply these filters successively in order to obtain filters with other frequency responses. Another possibility is to use a biquad 'filter recipe'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e. its first direct form consists of two quadratic polynomials in the fraction: $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 . z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
Both alternatives are explored by~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.

@@

\begin{enumerate}
  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of Figure~\ref{fig:iir}. The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs. The coefficients $a_0$ and $b_1$ of the IIR filter are given by $x \in [e^{-\pi},1]$:
\begin{equation}\label{eq:passa-baixas}
 \begin{split}
 x & =e^{-2\pi f_c} \\
 a_0 & =  1-x \\
 b_1 & =  x
 \end{split}
\end{equation}
  \item High-pass filter with a simple pole, module of its frequency responses at the upper right corner of Figure~\ref{fig:iir}. The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of $x \in [e^{-\pi},1]$:
\begin{equation}\label{eq:passa-altas}
 \begin{split}
 x & =e^{-2\pi f_c} \\
 a_0 & =  \frac{x+1}{2} \\
 a_1 & =  -\frac{x+1}{2} \\
 b_1 & =  x
 \end{split}
\end{equation}
\item Notch filter. This filter is parametrized by a center frequency $f_c$ and bandwidth $bw$, both given as fraction of $f_s$, therefore $f,\; bw \in (0,0.5)$. Both frequencies $f_c \pm bw$ have $\approx 0.707$ of the amplitude, i.e. attenuation of $3dB$. The auxiliary variables $K$ and $R$ are defined as:
\begin{equation}\label{eq:varAux}
 \begin{split}
  R & = 1 - 3bw \\
  K & = \frac{1-2R\cos(2\pi f_c) + R^2}{2 - 2 \cos (2 \pi f_c)}
 \end{split}
\end{equation}
The band-pass filter in the lower left corner of Figure~\ref{fig:iir} has the following coefficients:
\begin{equation}\label{eq:passa-banda}
 \begin{split}
 a_0 & =  1 - K \\
 a_1 & =  2(K-R)\cos (2\pi f_c) \\
 a_2 & =  R^2-K \\
 b_1 & =  2R \cos (2\pi f_c) \\
 b_2 & =  -R^2
 \end{split}
\end{equation}
The coefficients of band-reject filter, depicted in the lower right of Figure~\ref{fig:iir}, are:
\begin{equation}\label{eq:rejeita-banda}
 \begin{split}
 a_0 & =  K \\
 a_1 & =  -2K\cos (2\pi f_c) \\
 a_2 & =  K \\
 b_1 & =  2R \cos (2\pi f_c) \\
 b_2 & =  -R^2
\end{split}
\end{equation}
\end{enumerate}

\subsection{Noise}\label{subsec:ruidos}
Sounds without an easily recognizable pitch are generally called noise~\cite{Lacerda}. They are important musical sounds, as noise is present in real notes, e.g. emitted by a violin or a piano. Furthermore, many percussion instruments does not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}. In electronic music, including electro-acoustic and dance genres, noise has diverse uses and frequently characterizes the music style~\cite{Cook}.

@@

Some noises with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors for many reasons.
Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic sequence side-by-side. All five noises were generated with the same phase for each component, making it straightforward to observe the contributions of different parts of the spectrum.

\begin{itemize}
 \item The white note has its name because its energy is distributed equally among all frequencies, such as the white color. It is possible to obtain white noise with the inverse transform of the following coefficients:
\begin{equation}\label{eq:branco}
 \begin{split}
 c_0 & =0 \;\text{,\quad to avoid bias} \\
 c_i & =e^{j.x}\;,\; x \; \text{random} \; \in \; [0,2\pi]\;,\; i \; \in \; \left[1, \, \frac{\Lambda}{2}-1\right] \\
 c_{\Lambda/2} & = 1 \; \text{, \; if $\Lambda$ even}\\ 
 c_i & = c_{\Lambda - i}^*\;,\;\; \text{for}\;  i \; > \;  \frac{\Lambda}{2}
 \end{split}
\end{equation}

@@

 \item The gray noise is defined as a white noise subject to one of the ISO-audible curves. Those curves are obtained by experiments and are imperative to obtain $\alpha_i$. An implementation of ISO 226, which is the last revision of these curves, is in the \massa\ toolbox~\cite{MASSA}.
\end{itemize}
